---
title: "Introduction to ClinSim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clinsim-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ClinSim)
```



## Overview

This vignette shows how to use `clinsim`. The `clinsim` package validates sample size calculations by simulating the actual test many times and checking whether the empirical power matches the target.

The package provides two main functions:

-   **`wpower()`** — sample size calculation and validation for the Wilcoxon-Mann-Whitney test
-   **`mean2power()`** — sample size validation for the two-sample mean comparison (t-test)

## wpower()

The nonparametric setup in `wpower()` is based on a simple location-shift model.

Let

$$
X = e, \qquad Y = \theta + e'
$$

where $e$ and $e'$ are independent random errors drawn from the same distribution $F$. The parameter $\theta$ represents the treatment effect (a shift in location).

Under this model,

$$
p_1 = P(Y > X)
$$

is the probability that an observation from the treatment group exceeds one from the control group. Two additional probabilities are used in the TrialSize sample size approximation:

$$
p_2 = P(Y > X_1, Y > X_2), \qquad
p_3 = P(Y_1 > X, Y_2 > X)
$$

where $X_1,X_2$ and $Y_1,Y_2$ are independent draws from the control and treatment distributions.

The function `wpower()` estimates these probabilities and then compares two sample size approaches:

-   **Noether (1987)** — depends only on $p_1$
-   **TrialSize-style approximation** — depends on $p_1,p_2,p_3$

The recommended sample sizes are then evaluated by simulating the Wilcoxon rank-sum test using `stats::wilcox.test()`.

### How to use
At minimum, provide `theta` (the location shift you want to detect):

```{r basic, eval = FALSE}
Example0 <- wpower(theta = 0.7)
print(Example0)
```

### Detailed output

```{r summary, eval = FALSE}
summary(Example0)
```

### Three ways to specify the probability parameters

The TrialSize formula needs three probability parameters: $p_1$, $p_2$, and $p_3 $. Noether only needs $p_1$. You have three options for how these are obtained:

**Estimate from simulation (default):** The package generates large samples under the assumed model and computes the proportions.

```{r est-sim, eval = FALSE}
Example1 <- wpower(theta = 0.7)
```

**Estimate from pilot data:** Provide control and treatment data from a pilot study.

```{r est-pilot, eval = FALSE}
pilot_x <- rnorm(50)
pilot_y <- rnorm(50, mean = 0.7)
Example2 <- wpower(theta = 0.7, pilot_x = pilot_x, pilot_y = pilot_y)
```

**Set directly:** Specify any or all of the parameters yourself. You can mix — for example, set $p_1$ but let $p_2$ and $p_3$ be estimated.

```{r est-direct, eval = FALSE}
Example3 <- wilpower(theta = 0.7, p1 = 0.7, p2 = 0.56, p3 = 0.56)
```

### Changing the error distribution

By default, normal errors are assumed. Supply any random number generator via `rdist`:

```{r dist, eval = FALSE}
# Uniform errors
Example4 <- wilpower(theta = 0.7, rdist = function(n) runif(n, -1, 1))
```


### Passing arguments to wilcox.test

Additional arguments pass through to `stats::wilcox.test()` via `...`:

```{r dots-wil, eval = FALSE}
# One-sided test
Example5 <- wilpower(theta = 0.7, alternative = "greater")

# Exact test
Example6 <- wilpower(theta = 0.7, exact = TRUE)

# With continuity correction
Example7 <- wilpower(theta = 0.7, correct = TRUE)
```

### Allocation

Change the allocation ratio $k = n_1 / n_2$:

```{r unequal, eval = FALSE}
Example8 <- wilpower(theta = 0.7, k = 2)
```

### Power/Sample curves

To see how both methods perform across a range of effect sizes:

```{r power-curve, eval = FALSE}
Example9 <- powerscan(theta_range = seq(0.3, 1.0, by = 0.1))

```


## Two-Sample Mean: mean2power()

You have computed a sample size for a two-sample t-test and want to verify via simulation that the empirical power matches the target.

### How to use

```{r basic-mean, eval = FALSE}
Example7 <- mean2power(alpha = 0.05, beta = 0.2, sigma = 1, k = 1, margin = 0.3)
Example$empirical_power
```
